Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-10-02 demystify.tex
updates

\paragraph{Adcom} After screening applications are reviewed by a PhD admission committee or \emph{adcom} that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At Mason, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review about 30 applications. Note that most large schools, including Mason, have separate adcoms for MS programs (\autoref{chap:ms}).

@@

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various types of faculty). This provides junior faculty the opportunities to recruit students. Some members are tenured faculty. The chair of the committee will likely be a senior faculty, but they will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty, e.g., I am often assigned to review applicants interested in software engineering.

@@

\paragraph{Why CS depts do not waive the application fee?}  Some universities do waive, e.g., Rice does for all PhD applicants and many universities for domestic students (\autoref{sec:domestic-students}), however most do not for international students.  In any case, the application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Also, most CS programs already receive many applications and would be overwhelmed if the application process were free.  Note that if you have financial difficulties, you can ask the department for a waiver, but this is typically only granted in exceptional cases.

@@

@@

\subsection{Why did you get rejected?}\label{sec:why-rejected}

\sectioninfo{You aim too high, are overqualified, or even because you applied to AI/ML, a super competitive field in recent years with many applicants.}

@@

\subsection{Timeline: Why are rejection letters sent so late?}\label{sec:late-rejection}
\sectioninfo{Why rejection letters are sent so late? Grad programs often wait for the accepted students to make their decisions, typically by April 15, before sending out rejection letters.}

@@

@@

One confusion students often have is whether they will get paid over the summer.  The answer is that it depends on the funding source.
First, remember that an academic year (AY), which consists of Fall and Spring semesters, is typically 9 months, so your stipend is for 9 months (and many places allow you to spread it over 12 months).

If your funding source is TA (\autoref{sec:ta}), you typically do not get paid over the summer as there are few courses.
Some CS departments offer some summer funding, but it is typically not guaranteed and might not be a lot. For example, at Mason, we offer summer funding for 1st-year PhD TA students. The amount over the 3-month summer is similar to their monthly stipend (i.e., their 9-month stipend divided by 3 for the 3 summer months).

For RA (\autoref{sec:ta}), it depends on your prof. and their funding. When writing grant proposals, profs. typically include summer funding for their students (\autoref{sec:ra-cost}). However, funding is never guaranteed, e.g., the prof might not get the grant.
For my students, I have been fortunate to have funding to support them over the summer. Over the 3-summer months, I typically pay them 1/3 of their 9-month stipend. I prioritize summer funding for my students because Mason has very good TA resources so they never have to worry about funding during the AY.

@@

\section{Increasing your admission chance}\label{sec:improve-your-chance}

\sectioninfo{Increasing admission chance by being unique and standing out.}

@@

\section{How are decisions made?}\label{sec:how-decisions}
\sectioninfo{Even if all adcom reviewers recommend acceptance, the application can still be rejected. Vice versa, if all reviewers think the application is weak, the student might still be admitted.}

\section{Do adcom members talk to each other?}\label{sec:adcom-discuss}
\sectioninfo{Sometimes adcom members discuss applicants, but in most cases they make independent decisions.}

\section{Do adcom members read all applications?}\label{sec:read-all}
\sectioninfo{Adcom members only review applications assigned to them (typically matching their expertise) and rarely get involved in other applications}

@@

%\autoref{tab:ranking} lists .

\chapter*{History and Acknowledgement}\label{sec:ack}

\paragraph{History} This document was conceived during a lunch with a faculty at Mason.  We talked about why Mason was not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \autoref{sec:selecting-ranking-schools}). We wish there were a way for international students to know about PhD programs in the US.

>>>>

\subsection*{Admission Committee (adcom)}\label{sec:adcom}

Your applications are reviewed by a PhD admission committee or \emph{adcom} that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At Mason, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review about 30 applications. Note that most large schools, including Mason, have separate adcoms for MS programs (\autoref{chap:ms}).

@@

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various types of faculty). This provides junior faculty the opportunities to recruit students. The chair of the committee will likely be a senior faculty, but they will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty, e.g., I am often assigned to review applicants interested in software engineering.

@@

@@

\subsubsection{Why CS depts do not waive the application fee?}

Some universities do waive, e.g., Rice does for all PhD applicants and many universities for domestic students (\autoref{sec:domestic-students}), however most do not for international students.  In any case, the application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

\begin{commentbox}{Vu:}
In my opinion, requiring applicants to pay the fee helps ensure their seriousness, as it filters out non-serious candidates. Also, most CS programs already receive many applications and would be overwhelmed if the application process were free.  Note that if you have financial difficulties, you can ask the department for a waiver, but this is typically only granted in exceptional cases.
\end{commentbox}
\subsection{How are decisions made?}\label{sec:how-decisions}
\subsectioninfo{Even if all adcom reviewers recommend acceptance, the application can still be rejected. Vice versa, if all reviewers think the application is weak, the student might still be admitted.}

\subsection{Do adcom members talk to each other?}\label{sec:adcom-discuss}
\subsectioninfo{Sometimes adcom members discuss applicants, but in most cases they make independent decisions.}

\subsection{Do adcom members read all applications?}\label{sec:read-all}
\subsectioninfo{Adcom members only review applications assigned to them (typically matching their expertise) and rarely get involved in other applications}

@@

\subsection{Why did you get rejected?}\label{sec:why-rejected}
\subsectioninfo{You aim too high, are overqualified, or even because you applied to AI/ML, a super competitive field in recent years with many applicants.}

@@

\subsection{Timeline: Why are rejection letters sent so late?}\label{sec:late-rejection}
\subsectioninfo{Grad programs often wait for the accepted students to make their decisions, typically by April 15, before sending out rejection letters.}

@@

\subsection{Increasing your admission chance}\label{sec:improve-your-chance}
\subsectioninfo{You can improve your profile by being unique and standing out.}

@@

One confusion students often have is whether they will get paid over the summer.  This depends on the funding source.
First, recall that an academic year (AY), which consists of Fall and Spring semesters, is typically 9 months, so your stipend is for 9 months (and many places allow you to spread it over 12 months).

If your funding source is TA (\autoref{sec:ta}), you typically do not get paid over the summer, which has few courses.
Some CS departments offer summer funding, but it is not guaranteed and might not be a lot. For example, at Mason, we offer summer funding for 1st-year PhD TA students. The amount over the 3-month summer is similar to their monthly stipend (i.e., their 9-month stipend divided by 3 for the 3 summer months).

For RA (\autoref{sec:ta}), it depends on your prof. and their funding. When writing grant proposals, profs. typically include summer funding for their students (\autoref{sec:ra-cost}). However, funding is never guaranteed, e.g., the prof might not get the grant.

\begin{commentbox}[Vu:]
For my students, I have been fortunate to have funding to support them over the summer. Over the 3-summer months, I typically pay them 1/3 of their 9-month stipend. I prioritize summer funding for my students because Mason has very good TA resources so they never have to worry about funding during the AY.
\end{commentbox}

@@

@@

@@

%\autoref{tab:ranking} lists .
\chapter{History and Acknowledgement}\label{sec:ack}
\paragraph{History} This document was conceived during a lunch with a faculty at Mason.  We talked about why Mason was not able to attract good Vietnamese and other international students, despite having a much stronger CS program than many schools that these students want to go to (part of the reason is described in \autoref{sec:selecting-ranking-schools}). We wish there were a way for international students to know about PhD programs in the US.
